Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-29 submission/manuscript.Rmd
abstract edits

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). While, loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, the mechanisms and consequences are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating the role of lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are typically decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene dysregualtion was sexually dimorphic, as female _Kdm5c_-KO EpiLCs were more prone to germline gene misexpression than knockout males. Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, CGI-free germline gene promoters lacking KDM5C, particularly late-stage spermatogenesis genes, can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Thus, ectopic germline transcriptional programs can progress in the background of typical _Kdm5c_-KO development, due to downstream activation by key drivers of germline gene expression.

>>>>

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are usually decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression. Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Thus, distinct suppressive mechanisms are recruited to different germline gene classes and germline transcriptional programs can mirror germ cell differentiation in the background of typical somatic tissue development.
